The Gamma: Data Exploration through Iterative Prompting

### 2020-09-11 wip-design/paper.tex
Design paper - related work and introduction

\documentclass{sigchi}
\usepackage{xspace}
\usepackage{fancyvrb}
\usepackage{inconsolata}
\usepackage{xcolor}
\usepackage{tikz}
\newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
    \draw (0,0) circle (1);
    \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
    \end{tikzpicture}}
\newcommand*{\priorityc}[1]{
\resizebox{0.8em}{0.8em}{
  {\protect\tikz{ \protect
    \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
    } }}}

\definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
\DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
%\newcommand{\kvd}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\kvd}[1]{\textbf{#1}}
\newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}
\newcommand{\tg}{The Gamma\xspace}

% Use this section to set the ACM copyright statement (e.g. for
% preprints).  Consult the conference website for the camera-ready
% copyright statement.

% Copyright
\CopyrightYear{2020}
%\setcopyright{acmcopyright}
\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}
% DOI
\doi{https://doi.org/10.1145/3313831.XXXXXXX}
% ISBN
\isbn{978-1-4503-6708-0/20/04}
%Conference
\conferenceinfo{UIST'20,}{October  20--23, 2020, Minneapolis, MN, USA}
%Price
\acmPrice{\$15.00}

% Use this command to override the default ACM copyright statement
% (e.g. for preprints).  Consult the conference website for the
% camera-ready copyright statement.

%% HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP --
%% Please note you need to make sure the copy for your specific
%% license is used here!
% \toappear{
% Permission to make digital or hard copies of all or part of this work
% for personal or classroom use is granted without fee provided that
% copies are not made or distributed for profit or commercial advantage
% and that copies bear this notice and the full citation on the first
% page. Copyrights for components of this work owned by others than ACM
% must be honored. Abstracting with credit is permitted. To copy
% otherwise, or republish, to post on servers or to redistribute to
% lists, requires prior specific permission and/or a fee. Request
% permissions from \href{mailto:Permissions@acm.org}{Permissions@acm.org}. \\
% \emph{UIST '20},  October 20--23, 2020, Minneapolis, MN, USA \\
% ACM xxx-x-xxxx-xxxx-x/xx/xx\ldots \$15.00 \\
% DOI: \url{http://dx.doi.org/xx.xxxx/xxxxxxx.xxxxxxx}
% }

% Arabic page numbers for submission.  Remove this line to eliminate
% page numbers for the camera ready copy
% \pagenumbering{arabic}

% Load basic packages
\usepackage{balance}       % to better equalize the last page
\usepackage{graphics}      % for EPS, load graphicx instead
\usepackage[T1]{fontenc}   % for umlauts and other diaeresis
\usepackage{txfonts}
\usepackage{mathptmx}
\usepackage[pdflang={en-US},pdftex]{hyperref}
\usepackage{color}
\usepackage{booktabs}
\usepackage{textcomp}

% Some optional stuff you might like/need.
\usepackage{microtype}        % Improved Tracking and Kerning
% \usepackage[all]{hypcap}    % Fixes bug in hyperref caption linking
\usepackage{ccicons}          % Cite your images correctly!
% \usepackage[utf8]{inputenc} % for a UTF8 editor only

% If you want to use todo notes, marginpars etc. during creation of
% your draft document, you have to enable the "chi_draft" option for
% the document class. To do this, change the very first line to:
% "\documentclass[chi_draft]{sigchi}". You can then place todo notes
% by using the "\todo{...}"  command. Make sure to disable the draft
% option again before submitting your final document.
\usepackage{todonotes}

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{The Gamma: Data Exploration through Iterative Prompting}
\def\plainauthor{First Author, Second Author, Third Author,
  Fourth Author, Fifth Author, Sixth Author}
\def\emptyauthor{}
\def\plainkeywords{Data exploration; End-user programming; Data journalism; Programming languages; Type providers}
\def\plaingeneralterms{Documentation, Standardization}

% llt: Define a global style for URLs, rather that the default one
\makeatletter
\def\url@leostyle{%
  \@ifundefined{selectfont}{
    \def\UrlFont{\sf}
  }{
    \def\UrlFont{\small\bf\ttfamily}
  }}
\makeatother
\urlstyle{leo}

% To make various LaTeX processors do the right thing with page size.
\def\pprw{8.5in}
\def\pprh{11in}
\special{papersize=\pprw,\pprh}
\setlength{\paperwidth}{\pprw}
\setlength{\paperheight}{\pprh}
\setlength{\pdfpagewidth}{\pprw}
\setlength{\pdfpageheight}{\pprh}

% Make sure hyperref comes last of your loaded packages, to give it a
% fighting chance of not being over-written, since its job is to
% redefine many LaTeX commands.
\definecolor{linkColor}{RGB}{6,125,233}
\hypersetup{%
  pdftitle={\plaintitle},
% Use \plainauthor for final version.
%  pdfauthor={\plainauthor},
  pdfauthor={\emptyauthor},
  pdfkeywords={\plainkeywords},
  pdfdisplaydoctitle=true, % For Accessibility
  bookmarksnumbered,
  pdfstartview={FitH},
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=black,
  urlcolor=linkColor,
  breaklinks=true,
  hypertexnames=false
}

% create a shortcut to typeset table headings
% \newcommand\tabhead[1]{\small\textbf{#1}}

% End of preamble. Here it comes the document.
\begin{document}

\title{\plaintitle}

\numberofauthors{1}
\author{%
  \alignauthor{Leave Authors Anonymous\\
    \affaddr{for Submission}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}
}

@@

\begin{abstract}
Governments, non-profit organizations and citizen initiatives publish increasing amounts of
data, but extracting insights from such data and presenting them to the public is hard.
First, data comes in a variety of formats that each requires a different tool. Second, many
data exploration tools do not reveal how a result was obtained, making it difficult to reproduce
the results and check how they were obtained.
%
We contribute The Gamma, a novel data exploration environment for non-experts. The Gamma is based
on a single interaction principle and using it results in transparent and reproducible scripts.
This allows transfer of knowledge from one data source to another and
learning from previously created data analyses. We evaluate the usability and learnability of
The Gamma through a user study on non-technical employees of a research institute.
%
We argue that the our approach allows journalists and the public to benefit from the rise
of open data, by making data exploration easier, more transparent and more reproducible.
\end{abstract}

% ACM Classfication
%
% \begin{CCSXML}
% <ccs2012>
% <concept>
% <concept_id>10003120.10003121</concept_id>
% <concept_desc>Human-centered computing~Human computer interaction (HCI)</concept_desc>
% <concept_significance>500</concept_significance>
% </concept>
% <concept>
% <concept_id>10003120.10003121.10003125.10011752</concept_id>
% <concept_desc>Human-centered computing~Haptic devices</concept_desc>
% <concept_significance>300</concept_significance>
% </concept>
% <concept>
% <concept_id>10003120.10003121.10003122.10003334</concept_id>
% <concept_desc>Human-centered computing~User studies</concept_desc>
% <concept_significance>100</concept_significance>
% </concept>
% </ccs2012>
% \end{CCSXML}
%
% \ccsdesc[500]{Human-centered computing~Human computer interaction (HCI)}
% \ccsdesc[300]{Human-centered computing~Haptic devices}
% \ccsdesc[100]{Human-centered computing~User studies}

% Author Keywords
\keywords{\plainkeywords}

% Print the classficiation codes
% \printccsdesc
% Please use the 2012 Classifiers and see this link to embed them in the text: \url{https://dl.acm.org/ccs/ccs_flat.cfm}

\begin{figure}
\includegraphics[width=1\columnwidth]{figures/thegamma-annot}
\caption{Teams with the greatest number of gold medals from Rio 2016
Olympics with a reproducible The Gamma script (1), an auto-complete prompt offering ways
of sorting the data (2) and instant preview (3).}
\label{fig:thegamma}
\vspace{-0.5em}
\end{figure}

\section{Introduction}
Data science has more capabilities to help us understand the world than ever before, yet at the
same time post-truth politics and increasing public distrust in statistics makes data-driven insights
increasingly less relevant in public discourse~\cite{howstatslost}. This should perhaps not be a surpirse.
Journalists can access increasing amounts of data, but producing engaging and transparent data-driven
reports that are easy to interpret is expensive and requires expert programming skills \cite{ddj}.

The design of a data exploration tool for journalists poses a unique mix of challenges. First, the
tool needs to be easy to learn for end-users working under tight deadlines. Second, it needs to
support a wide range of data sources in a way where the expertise gained when working with one data
source is relevant for other data sources. Third, the resulting data-driven insights need to be
transparent, allowing the readers to verify the claims and learn how to reproduce the work.

We present The Gamma, a text-based data exploration environment for non-experts. The Gamma
is based on a single interaction principle, which provides uniform
access to a range of data sources including data tables, graph databases and data cubes.
An anlysis created in The Gamma is a transparent script that can be followed to reproduce the
result from scratch. This allows learning from existing analyses and encourages readers
to engage with the results. Our key contributions are:

\begin{itemize}
\item We identify the design requirements for a data exploration tool for journalists
  (Section~\ref{sec:motivation}) and follow those to build a novel programming environment
  The Gamma (Section~\ref{sec:overview}).

\item We introduce \emph{iterative prompting} (Section~\ref{sec:design}),
  an interaction design principle that can be used to complete a variety of programming
  tasks in a uniform way that allows transfer of knowledge between different tasks.

\item We show how to use the iterative prompting principle for querying of distinct data
  sources including data tables, graph databases and data cubes (Section~\ref{sec:implementation}).

\item We discuss a number of case studies (Section~\ref{sec:cases}) and
  conduct a user study to evaluate the usability of The Gamma and the extent to which users can,
  (i) learn from examples and (ii)~transfer knowledge between tasks (Section~\ref{sec:study}).
\end{itemize}

The Gamma is available as open-source at \href{http://thegamma.net}{\small\bf\ttfamily thegamma.net}.

@@

The Gamma brings together recent programming language innovations for working with data
\cite{inforich}, work on making programming accessible to non-programmers \cite{enduser,smallmatter}
and work on tools for journalists such as Idyll \cite{idyll}. Related work includes a range of
visual and programmatic tools used by data scientists, work that aims to simplify programming
or explore alternative paradigms and various predictive tools that assist users in completing
programming or data exploration tasks.

\subsection{Data science tools}
Data exploration is typically done using spreadsheets, comprehensive visual tools or
via code using notebooks. Notebooks such as Jupyter \cite{jupyter} are also popular with technically
skilled journalists as they allow combining code with commentary and visual outputs. However,
notebooks can lead to messy hard to reproduce code \cite{messes,wrattler}.

Visual data analytics tools \cite{control,vizdom} such as Tableau \cite{tableau} do not require
programming skills, but require mastering a range of different interactions. They also make it
hard to see what data transformations have been done. Several systems for data exploration,
e.g.~Potter's Wheel and Wrangler \cite{potter,wrangler}, and visualization, e.g.~Lyra \cite{lyra},
provide a GUI, but record a transparent script that can be seen and modified by the user.

\subsection{Easier programming tools}
We aim to design a programming environment that is easy to use and learn. Many approaches to this
goal have been tried. Victor \cite{learnable,principle} introduced design principles that
inspired many to build live programming systems \cite{review,liveroad,lighttable} that give
immediate feedaback to help programmers understand how code relates to output and
exploratory systems \cite{variolite,exploratory} that assist with completing open-ended tasks.

To avoid difficulties with editing code as text, some systems use structured editors~\cite{structure-based,livenut,lamdu}.
In Subtext \cite{subtext,directprog} the language itself is co-designed with the editor to make
the interactions with code more natural. Finally, programming can be made easier by designing
high-level declarative abstractions. Tea does so for statistical analyses \cite{tea} and
Satyanarayan et al. \cite{interactionviz,vegalite} does so for interactive data visualizations.

The Gamma is live in that it gives an instant preview of the results. It is text-based,
although it is syntactically simple. The code written in The Gamma is high-level and so work
on declarative abstractions is a valuable inspiration.

\subsection{Code completion and assistants}
A key component in The Gamma design is the use of auto-complete for offering possible
operations. Code completion has been pioneerd by Kaiser~\cite{assistants}, but our work is more
directly inspired by the work on type providers \cite{inforich,fsdata,dotdriven}, which utilize
code completion to integrate external data into a statically typed programming language.

When using type providers, completions are offered as the user types code. However, completion
tools can also recmmend scripts when users begin interacting with data \cite{predictive,proactive}.
Alternatively, they can also be guided by natural language as for example Eviza~\cite{eviza},
which uses natural language for data exploration. CodeMend~\cite{codemend} uses natural language
for more general code edits, but uses data visualization as a case study.
Finally, data entry interfaces based on completion such as Dasher \cite{dasher} utilize an
interaction principle similar to our iterative prompting.

\subsection{Programming without writing code}

A number of tools for data exploration use alternative approaches to programming where
the user does not write code. In programming by example \cite{byexample}, the user gives
examples of desired results. In data science, this approach has been used for example for
specifying data transformations in spreadsheets and data extraction \cite{spreadsheetpbe,flashextract}.

In direct manipulation \cite{direct}, a program is specified by directly interacting with the
output. This has been used in the visual domain \cite{sketchnsketch}, but also for data querying
\cite{dynamicq,vlang}. The VQE language~\cite{visage} also considers how to allow code reuse and
modification in this context. Direct manipulation can also support data exploration by letting
users partially edit queries,~e.g. by changing quantifiers as in DataPlay~\cite{dataplay}.

@@

\tg is a text-based programming environment that allows non-experts create simple data exploration
scripts using a single interaction principle -- choosing an item from an auto-complete list.
It supports a range of data sources including tabular data, graph data and data cubes.

@@

@@

As illustrated in Figure~\ref{fig:cubetp}, the provider allows uses to select a data series
from the data cube. By choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
we restrict the cube to a two-dimensional plane. We then choose an indicator category
\ikvd{\textquotesingle Climate Change\textquotesingle} and a specific indicator
\ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}, obtaining a time series with
years as keys and emission data as values. The World Bank provider also supports filtering
by year and indicator.

@@

Our second type provider allows users to construct queries to explore data in tabular formats
such as CSV files. It is based on the theory developed Petricek~\cite{dotdriven}. Unlike the
data cube provider, the provider for tabular data does not just allow selecting a subset of the
data, but it can be used to construct SQL-like query. Consider the example from Figure~\ref{fig:thegamma}:

@@

\paragraph{RQ1: Can non-programmers explore data with The Gamma?}
We give participants one of four simeple data exploration tasks and see if they can complete the
task using The Gamma and, possibly, how much assistance they need.

@@

\paragraph{RQ3: Can users learn from just code samples?}
Our second hypothesis is that users can learn through precolation, i.e.~by looking at the source
code of published analyses. In one of our tasks, participants are given a code sample, but only a
minimal explanation of The Gamma environment.

@@

% REFERENCES FORMAT
% References must be the same font size as other body text.
\bibliographystyle{SIGCHI-Reference-Format}
\bibliography{paper}

>>>>

\documentclass[manuscript,review,anonymous]{acmart}
\usepackage{hyperref}

% NOTE: Good things
%  - what is iterative prompting? "Like auto-complete but less limited" - hmm, maybe
%  - sell more main idea: can do much with very limited set of commands / via auto-complete
%  - paper is convincing in that this works for SQL-like queries (maybe not general purpose, but this is good)
%  - journalism is a good focus / but needs to deemphasize this a bit because of the study
%  - #1 Limited language for data exploration (what is included / excluded / essential / could be added)
%  - #2 Iterative prompting - what's new? people can get to all programs via '.' and up/down/enter.

% NOTE: Random
%  - How choices were made?
%  - Terminology: data analysis / data exploration / data visualization - what is the goal of the system? (fair enough)
%    (it's not about writing charts - just about getting the right data - we have limited charting functionality,
%     one could integrate with Vega, or do future work based on iterative prompting)
%  - Technical details - you can only create valid programs!! (it's contextual)
%  - Minor: what was "demo" at the start of a session
%  - Table 1 vs. Table 2 numbers of participants
%  - What data quality is needed to make this work?
%  - "Magic Escalator of Knowledge" - remove bs?
%  - Clarify how we support 3D cubes (no unified API); mention "done by intern in X weeks"

% TODO Study improvements
% - The questionaire needs a bit more introduction when you present  the study.
% - When looking at the data for the questionaire, it seems somewhat inconclusive
%    as all the values lie around 2.5.  I think this warrants a bit more discussion.
% - Connected to this, I would recommend to highlight the successes of the study a bit more in the
%     discussion and  presentation of the results. You actually got non-programmers to do
%     sophisticated queries! That you didn't get support for all RQs is perhaps more a result of being quite  ambitious!

% TODO
%   - SAY SOMEWHERE: choose function *and* specify arguments via auto-complete
%   - ADD: table of all available operations? (shorten introduction)

% NOTE: Contribution
%  - What is the exact goal? (so that it can be compared with R, Python, Vega-Lite, Tableau)
%  - How the design goals were generated? (not by talking to journalists)
%  - "notebooks are messy" - not arguing we can fix that

% NOTE: Motivation
% - what if you could write the whole program using auto-complete?

% extend autocomplete into a full-blown interactive programming mechanism
% and show that it is viable for small programming tasks like data exploration
%
% --- remove some of the motivation (makes ppl ask bad questions)

%
% Iterative prompting - should we call it expression completion
% (is it more general result, or is it a new thing?)
% -> emphasize the why this is called differently paragraph
%
% Explain type providers a bit more (we say what it does, but not much about what
% it is - give dynamic typing in a static language)
%
% The 'then' operation was mysterious - can we add one line explanation to the operations?
% (yes we can and we should totally do that)

% \usepackage{xspace}
% \usepackage{fancyvrb}
% \usepackage{inconsolata}
% \usepackage{xcolor}
% \usepackage{tikz}
% \newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
%     \draw (0,0) circle (1);
%     \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
%     \end{tikzpicture}}
% \newcommand*{\priorityc}[1]{
% \resizebox{0.8em}{0.8em}{
%   {\protect\tikz{ \protect
%     \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
%     } }}}
%
% \definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
% \DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
% %\newcommand{\kvd}[1]{\textcolor{kvdclr}{#1}}
% \newcommand{\kvd}[1]{\textbf{#1}}
% \newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}
% \newcommand{\tg}{The Gamma\xspace}
% \usepackage{balance}       % to better equalize the last page
% \usepackage{graphics}      % for EPS, load graphicx instead
% \usepackage[T1]{fontenc}   % for umlauts and other diaeresis
% \usepackage{txfonts}
% \usepackage{mathptmx}
% \usepackage[pdflang={en-US},pdftex]{hyperref}
% \usepackage{color}
% \usepackage{booktabs}
% \usepackage{textcomp}

\setcopyright{acmwhatever}
\copyrightyear{2020}
\acmYear{2020}
%\acmDOI{10.1145/1122445.1122456}
\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
%\acmISBN{978-1-4503-XXXX-X/18/06}

\begin{document}
\title{The Gamma: Data Exploration through Iterative Prompting}

\author{Tomas Petricek}
\email{t.petricek@kent.ac.uk}
\affiliation{%
  \institution{University of Kent}
  \city{Canterbury, UK}
}
%\renewcommand{\shortauthors}{Trovato and Tobin, et al.}

\begin{abstract}
  Governments, non-profit organizations and citizen initiatives publish increasing amounts of
  data, but extracting insights from such data and presenting them to the public is hard.
  First, data comes in a variety of formats that each requires a different tool. Second, many
  data exploration tools do not reveal how a result was obtained, making it difficult to reproduce
  the results and check how they were obtained.
  %
  We contribute The Gamma, a novel data exploration environment for non-experts. The Gamma is based
  on a single interaction principle and using it results in transparent and reproducible scripts.
  This allows transfer of knowledge from one data source to another and
  learning from previously created data analyses. We evaluate the usability and learnability of
  The Gamma through a user study on non-technical employees of a research institute.
  %
  We argue that our approach allows journalists and the public to benefit from the rise
  of open data, by making data exploration easier, more transparent and more reproducible.

\end{abstract}

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003120.10003121.10003124</concept_id>
       <concept_desc>Human-centered computing~Interaction paradigms</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011050.10011017</concept_id>
       <concept_desc>Software and its engineering~Domain specific languages</concept_desc>
       <concept_significance>300</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011066.10011069</concept_id>
       <concept_desc>Software and its engineering~Integrated and visual development environments</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Human-centered computing~Interaction paradigms}
\ccsdesc[500]{Software and its engineering~Integrated and visual development environments}
\ccsdesc[300]{Software and its engineering~Domain specific languages}
\keywords{data exploration; end-user programming; data journalism; programming languages; type providers}

@@

\begin{figure}
\includegraphics[width=.54\columnwidth]{figures/thegamma-annot}
\caption{Obtaining teams with the greatest number of gold medals from Rio 2016
Olympics with a reproducible The Gamma script (1), contextual iterative prompting mechainsm
offering ways of sorting the data (2) and an instant preview of results (3).}
\label{fig:thegamma}
\end{figure}

\section{Introduction}
Data science has more capabilities to help us understand the world than ever before, yet at the
same time post-truth politics and increasing public distrust in statistics makes data-driven insights
increasingly less relevant in public discourse~\cite{howstatslost}. To reverse this trend, we
need tools that let non-experts, including journalists and other information-literate citizens,
produce transparent, engaging data analyses that are easy to interpret without requiring expert
programming skill~\cite{ddj}. The design of such data exploration tool poses a unique mix of chellenges.
First, the tool needs to have a very low barrier to entry. Second, it needs to support a wide
range of data sources in a uniform way. Third, the resulting data analyses need to assist
readers in learning how to reproduce the work and verify the claims it makes.

We contribut The Gamma, a text-based data exploration environment for non-experts. The Gamma
is based on a single, easy to understand interaction principle and provides a uniform
access to a range of data sources including data tables, graph databases and data cubes.
The resulting analysis is a transparent script that can be followed to reproduce the
result from scratch. This allows learning from existing analyses and encourages readers
to engage with data.

\subsubsection*{Iterative prompting}
The key idea in The Gamma, which we term the \emph{iterative prompting} principle, is that all
valid data exploration scripts can be constructed by repeatedly choosing from a list
of offered options. This way, non-programmers can write entire scripts through auto-complete,
without learning a programming language, but they still produce transparent and reproducible
source code. In other words, iterative prompting turns auto-complete from a programmer assistance
tool into a non-expert programming mechanism. A crucial feature is that iterative prompting
only offer operations that are valid in a given context and that it offer all such operations;
it is both correct and complete.

\subsubsection*{Date exploration}
The Gamma focuses on data exploration of the kind illustrated in Figure~\ref{fig:thegamma}.
The user accesses data available in a structured format. They make several experiments to find an
interesting way of looking at the data, e.g.~by applying different aggregations or filters. They
may choose to view the results as a table or a basic chart before publishing their analysis. The
Gamma makes such simple programming with data simple enough for non-experts. Scraping and cleaning
of messy data or building rich data visualizations is outside of the scope of this paper, but
exposing those using the iterative prompting approach is an interesting future challenge.

\subsubsection*{Contributions}
The Gamma is available (non-anonymously) at \href{http://thegamma.net}{\textcolor{blue}{thegamma.net}},
both as a JavaScript library and a hosted data exploration service. In this paper, we describe and
evaluate the design principles behind the project:

\begin{itemize}
\item We introduce the iterative prompting principle in The Gamma (Section~\ref{sec:overview})
  and show that it can be used for querying of distinct data sources including data tables, graph
  databases and data cubes (Section~\ref{sec:implementation}).

\item We show that iterative prompting is a complete and correct program construction method and
  consider how it lowers the barrier to entry and allows learning through examples, without
  expert guidance (Section~\ref{sec:design}).

\item We evaluate the system through a number of case studies (Section~\ref{sec:cases})
  and a user study (Section~\ref{sec:study}). Our study shows that non-programmers can use The Gamma
  to construct non-trivial data queries and ascertains the extent to which users can,
  (i) learn from examples and (ii)~transfer knowledge between tasks.
\end{itemize}

@@

The key contribution of our work is that it develops a new, fundamentally different, way of using the
established auto-completion mechanism. Unlike most past work dating back to Kaiser~\cite{assistants},
we do not view it as a programmer assistance tool, but as an interaction mechanism through which
non-experts can create entire programs. We build on recent research on information-rich
programming \cite{inforich} and aim to make those advances available to non-programmers
\cite{enduser,smallmatter}, in the context of data exploration as done, most notably, by
journalists \cite{ddj}. Our work features a novel combination of characteristics in that
the iterative prompting we develop (i) is centered around editing and understanding of program code,
(ii) its conceptual complexity is reduced to a single basic kind of interaction, yet (iii) it is
correct and complete in that it can be used to construct all meaningful programs.

\subsubsection*{Code completion for data science}

A key component in The Gamma is the use of auto-complete for offering possible operations.
Our work follows type providers \cite{inforich,fsdata}, which integrate external data into a
static type system of F\#, allowing the use of auto-completion; for querying data tables, we utilize
the theory developed by Petricek \cite{dotdriven}. The key difference in our work is that The Gamma
can be used without a programming language expertise.

Most similar to our approach are tools that recommend scripts when users begin interacting
with data. Those based on machine learning code completion for domain-specific languages \cite{predictive,proactive}
differ in that they do not guarantee completeness, i.e.~the user cannot create all possible
scripts. Approaches based on natural language can effectively support data exploration
or visualization \cite{eviza,codemend}, but hide the structure of the underlying language and
require more than just selecting options. Conversational agents \cite{iris} improve on such work
in that they can provide more guidance. Code completion based on machine learning or statistical
methods \cite{mlcomplete,statcomplete} also exists for general-purpose programming languages used
by data scientists such as Python \cite{pythia}, providing assistance to expert programmers.
Finally, DS.js \cite{dsjs} is interesting in that it enables querying of data on the
web; it uses JavaScript with rich contextual code completion.

\subsubsection*{Notebooks and business intelligence tools}

Notebooks such as Jupyter \cite{jupyter}, which allow combining source code with commentary and
visual outputs, are widely used by data scientists, but require expert programming skills.
The Gamma targets non-experts, but could be easily integrated with a multi-language notebook system
such as Wrattler \cite{wrattler}.

Spreadsheets and business intelligence tools \cite{tableau,powerbi} do
not involve programming, but require mastering a complex GUI. This is also the case
for other visual data analytics tools \cite{control,vizdom}. In contrast, The Gamma is
based on a single kind of interaction. Several visual systems \cite{potter,wrangler,lyra} record
interactions with the GUI as a script that can be seen and modified by the user.
Unlike in The Gamma, the source code does not guide the user in learning how to use the system.

\subsubsection*{Easier programming tools}
We aim to build an easy to use and learn programming system. Many approaches to this
goal have been tried. Victor \cite{learnable,principle} introduced design principles that
inspired many to build live programming systems \cite{review,liveroad,lighttable} that give
immediate feedback to help programmers understand how code relates to output and
exploratory systems \cite{variolite,exploratory} that assist with completing open-ended tasks.
A system combining textual language with visualization also exists for graph querying \cite{guess}.
To avoid difficulties with editing code as text, some systems use structured editors~\cite{structure-based,livenut,lamdu}.
In Subtext \cite{subtext,directprog} the language itself is co-designed with the editor to make
the interactions with code more natural. The Gamma is live in that our editor gives an instant
preview of the results.

Many systems simplify programming by designing high-level declarative abstractions,
e.g.~for interactive news articles \cite{idyll}, statistical analyses \cite{tea}
or interactive data visualization \cite{interactionviz,vegalite}. The Gamma uses
high-level abstractions for data querying, but designing high-level iterative prompting
abstractions for other tasks remains future work.

\subsubsection*{Programming without writing code}
There are two main approaches to programming where
the user does not write code. In programming by example \cite{byexample}, the user gives
examples of desired results. This has been used, e.g.~for specifying data transformations
in spreadsheets and data extraction \cite{spreadsheetpbe,flashextract}.
In direct manipulation \cite{direct}, a program is specified by directly interacting with the
output. This has been used in the visual domain \cite{sketchnsketch}, but also for data querying
\cite{dynamicq,vlang}. The VQE language~\cite{visage} also considers how to allow code reuse and
modification in this context. Direct manipulation can also support data exploration by letting
users partially edit queries,~e.g. by changing quantifiers as in DataPlay~\cite{dataplay}.

\subsubsection*{Guestures and data entry}
Although our focus is on program construction, our work can be positioned in the
broader context of input methods. Akin to Dasher \cite{dasher}, our system provides a way of
navigating through a complete space of options, while on-screen feedforward \cite{octopocus} allows
efficient selection in guesture-based interfaces. Those provide compelling alternatives to
auto-completion menus, although the efficiency of input methods is not an issue in programming.

\section{Overview}

design motivated by exploratory question - what if we could program with autocomplete
no talking to journalists hence shorter paper

maybe talk about "design space"

\newpage

@@

\tg is a text-based programming environment that allows non-experts to create simple data exploration
scripts using a single interaction principle -- choosing an item from an auto-complete list.
It supports a range of data sources including tabular data, graph data and data cubes.

@@

\subsection{Correct and complete}
any program can be created by autocomplete, all programs created by autocomplete are correct

@@

As illustrated in Figure~\ref{fig:cubetp}, the provider allows users to select a data series
from the data cube. By choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
we restrict the cube to a two-dimensional plane. We then choose an indicator category
\ikvd{\textquotesingle Climate Change\textquotesingle} and a specific indicator
\ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}, obtaining a time series with
years as keys and emission data as values. The World Bank provider also supports filtering
by year and indicator.

@@

Our second type provider allows users to construct queries to explore data in tabular formats
such as CSV files. It is based on the theory developed Petricek~\cite{dotdriven}. Unlike the  % grammar!
data cube provider, the provider for tabular data does not just allow selecting a subset of the
data, but it can be used to construct SQL-like query. Consider the example from Figure~\ref{fig:thegamma}:

@@

\paragraph{RQ1: Can non-programmers explore data with The Gamma?}
We give participants one of four simple data exploration tasks and see if they can complete the
task using The Gamma and, possibly, how much assistance they need.

@@

\paragraph{RQ3: Can users learn from just code samples?}
Our second hypothesis is that users can learn through percolation, i.e.~by looking at the source
code of published analyses. In one of our tasks, participants are given a code sample, but only a
minimal explanation of The Gamma environment.

@@

% REFERENCES FORMAT
% References must be the same font size as other body text.
\bibliographystyle{ACM-Reference-Format}
\bibliography{paper}
